---
title: "R Notebook"
output: html_notebook
creator: emiliomora92@gmail.com 
---

This code changes the rawData_wayPointsCompletos.csv to data suitable to analysis. I explain the different procedures

```{r}
library("methods")
library("xml2")
library(tidyverse)
library(sf) #gestion de archivos de forma

```

```{r}
WP_COSECHA <- read.csv("./rawData_wayPointsCOMPLETOS.csv", stringsAsFactors = FALSE)
```

This chunks extract latitude and longitude from "position" column and convert it to decimal values, using common expression

```{r}
#https://stringr.tidyverse.org/articles/regular-expressions.html
WP_COSECHA <- WP_COSECHA %>% separate(posicion, c("lat1", "lat2", "long1", "long2"), sep=" ") %>%
  unite("lon", long1:long2, sep="") %>%
  unite("lat", lat1:lat2, sep="")

#here we extract the value of long and lat without the N. So it ask for digits until in reaches ? (in this case°)
# and then it add the values after the ° and divide this by 60 to have it all in degrees. 
WP_COSECHA$lon <- as.numeric(str_extract(WP_COSECHA$lon, "[:digit:]+?(?=°)"))+ 
  as.numeric(str_extract(WP_COSECHA$lon, "(?<=°).+?(?=')"))/60 

WP_COSECHA$lat <- as.numeric(str_extract(WP_COSECHA$lat, "[:digit:]+?(?=°)"))+ 
  as.numeric(str_extract(WP_COSECHA$lat, "(?<=°).+?(?=')"))/60
```

Here we take the date, we divide it in date and hour, then the hour is transformed to minutes.
```{r}
WP_COSECHA$arboles <- 1
WP_COSECHA<- WP_COSECHA %>% separate(Fecha.de.creación.mod, c("fecha", "hora"), sep = " ")
WP_COSECHA$Fecha.de.modificación <- NULL
WP_COSECHA$hora <- as.character(WP_COSECHA$hora)
WP_COSECHA$hora <- as.numeric(WP_COSECHA$hora)
WP_COSECHA$tiempo <- (as.integer(WP_COSECHA$hora)*60)+((WP_COSECHA$hora-as.integer(WP_COSECHA$hora))*100) #to have it all in minutes  
```


We create 2 vector to then sustract them and have the delta time- The delta time T2-T1 will be in T1 row. 
This is helpful to remove the "pause times" because these are defined as T4 (reinitit, tree_X) - T3 (pause, NA) and will be in the (T3, pause, NA, delta_pauseTime) and we will remove that row later (see full explanation in paper)
```{r}

horaA <-WP_COSECHA$tiempo
horaB <- horaA
horaA[length(horaA)+1]<- NA
horaA<- horaA[2:length(horaA)]  #
delta<- horaA-horaB
WP_COSECHA$delta <- delta

WP_COSECHA_sP <- WP_COSECHA %>% filter(!(punto=="NA"))  #

#we remove person, sex and age
WP_COSECHA_sP <- select(WP_COSECHA_sP, -5, -6, -7)
WP_COSECHA_sP$zona[WP_COSECHA_sP$zona == "falta"] <- "zonaBaja"


colnames(WP_COSECHA_sP)[1]<- "ID_REC"
write.csv(WP_COSECHA_sP, "./cleanData_wayPointsCoffee_latlon.csv", row.names = FALSE)
```


Now we have a clean data frame. We use this csv to create, using latlong, the X_UTM and Y_UTM coordinate. The Lon Lat were in WGS84 and UTM region is 15N

```{r}
# Transformar csv en objeto espacial
SF_Cafe <- WP_COSECHA_sP %>% 
  st_as_sf(coords = c("lon", #nombre de la columna con la longitud
                      "lat"), #nombre de la columna con la latitud
           crs = "EPSG:4326") #clave del datum WGS84

# Reproyectar SF de WGS84 a UTM 
SF_Cafe_UTM <- SF_Cafe %>% 
  st_transform(crs = "EPSG:32615") #clave de la zona UTM 15N

# Crear columnas con las coordenadas UTM
SF_Cafe_UTM <- SF_Cafe_UTM %>% 
  mutate(x_UTM = st_coordinates(.)[,1],
         y_UTM = st_coordinates(.)[,2])

WP_COSECHA_sP$x_UTM <- SF_Cafe_UTM$x_UTM
WP_COSECHA_sP$y_UTM <- SF_Cafe_UTM$y_UTM
```


Now we send this to the data directory for analysis 
```{r}
write.csv(WP_COSECHA_sP, "../distributionAnalisis/data/cleanData_wayPointsCoffee_UTM.csv", row.names = FALSE)
```




